---
title: "Lecture 12: Advanced data manipulation: joins, pivoting"
---

```{r}
#| label: setup

suppressPackageStartupMessages(library(tidyverse))
```

This lecture is the most important one we will cover on R. Most everyting else we saw in this course can be done with Excel. This, on the other hand, is something that is very difficult to do in Excel, and is one of the main reasons why we use R (or other programming languages) for data manipulation.

# Joins

Our problem: we have two datasets; one with information about students (e.g., name, age, grade), and another with information about their test scores (e.g., name, test score). We want to combine these two datasets into one dataset that contains all the information about the students and their test scores.

This is a very common problem in data analysis. The action of comining two dataets is called a **join**. In other programming languages (e.g., Stata), this is also called a **merge**. 

## Example 1: a one-to-one join

The notion of one-to-one join is conceptual, rather than technical. It means that there is a one-to-one relationship between the two datasets. In other words, each row in one dataset corresponds to exactly one row in the other dataset. In our example, each student took one test, so there is a one-to-one relationship between the students and their test scores.

```{r}
#| label: student-data

students <- tibble(
  id = c(1, 2, 3),
  name = c("Alice", "Bob", "Charlie"),
  age = c(20, 21, 22)
)
```

```{r}
#| label: tests-data-1

tests <- tibble(
  id = c(1, 2, 3),
  name = c("Alice", "Bob", "Charlie"),
  test_score = c(90, 85, 95)
)

# DOING THE JOIN
inner_join(students, tests, by = "id")
```

### Good practice 1: always specify the by argument when performing a join.

```{r}
#| label: tests-data-1-no-by

# DOING THE JOIN
# you should always specify the variable(s) that you want to join on, using the by argument
# if you don't specify the by argument, the join will be performed on all variables
# that have the same name in both datasets
inner_join(students, tests, by = "id")

# look at the output: the join was performed using the "id"
# variable, because that's the only variable that has the same name in both datasets
inner_join(students, tests)

# now let's look at a problematic example
students_pb <- students |>
  mutate(
    # this column is meant to reflect the student's overall letter grade
    letter_grade = c("A", "B", "C")
  )
tests_pb <- tests |>
  mutate(
    # while this column is meant to reflect the student's letter grade in my class
    # everyone got a D in my class
    letter_grade = "D"
  )

# this works fine:
# we're joining by the id variable; R creates 2 letter grade variables, one for each dataset
inner_join(students_pb, tests_pb, by = "id")

# this creates a problem:
# we're joining by id and letter_grade
# there are no matching rows
# id 1 has letter grade A in students_pb and D in tests_pb -> so no match for id 1
# id 2 has letter grade B in students_pb and D in tests_pb -> so no match for id 2,
# ... and so on
# -> so the result is empty

# bottom line: ALWAYS SPECIFY THE BY ARGUMENT WHEN PERFORMING A JOIN
# this will help you avoid unpleasant surprises like this one
inner_join(students_pb, tests_pb)
```

### Good practice 2: always join on a unique, numeric identifier

Remember that in R, `"Alice"` and `"alice "` are different values. If we join on the name variable, we may run into problems if there are typos in the names, or if there are multiple students with the same name. This is why it is a good practice to always join on a unique, numeric identifier (e.g., id).


## Example 2: a one-to-many join

Another type of relationship is a one-to-many relationship. In other words, each row in one dataset corresponds to multiple rows in the other dataset. In this example, each student took multiple tests, so there is a one-to-many relationship between the students and their test scores. There is a final type of relationship, called many-to-many, but we will not cover it in this lecture. Technically, whether the relationship is one-to-one, one-to-many, or many-to-many does not affect how we perform the join. However, it does affect how we interpret the results of the join.

```{r}
#| label: tests-data-2

tests <- tibble(
  id = c(1, 1, 2, 2, 3, 3),
  test_score = c(90, 85, 95, 80, 10, 20)
)

# you write joins for many-to-many relationships work just as for one-to-one relationships,
# but the output will have more rows
inner_join(students, tests, by = "id")
```

## Diagnosing problems with joins

There may be problems with the data that prevent us from performing the join correctly. For example, there may be duplicate rows in one of the datasets, or there may be missing values. It is important to diagnose these problems before performing the join, because they can lead to incorrect results. I like to think about this in the following way: 

1. What is relationship we expect to see between the two datasets? (one-to-one, one-to-many, many-to-many)? 
2. Do we expect to see missing data? 
3. How does this translate into 
    - the number of rows we expect to see in the joined dataset?
    - the number of rows we expect to see in each group of the joined dataset? 

Let's go back to our initial example: one grade per student.  

1. We expect to see a one-to-one relationship between the students and their test scores.
2. We do not expect to see missing data: all students should have exactly one grade.
3. We expect that 
    - the joined dataset will have 3 rows (one for each student)
    - each group (i.e., each student) will have 1 row (one grade per student)

Now, consider the messy dataset below. We deliberately introduced some problems:

- Student 2 (Bob) is missing (he didn't take the test, thus has no grade). 
- Student 3 (Charlie) took the test twice, thus has two grades.
- Student 4 took the test, but is not in the students dataset (maybe he is a new student who just joined the class, or maybe he is a student from another class who took the test by mistake).

```{r}
#| label: tests-data-3

tests <- tibble(
  id = c(1, 3, 3, 4),
  test_score = c(90, 85, 95, 10)
)

result <- inner_join(students, tests, by = "id")

# let's look at the output: it has 3 rows, as expected, but not for the reasons we expected.
# student 2 is gone, and student 3 has 2 rows.
result

# check that there are no duplicate rows in the result
sanity_check_1 <- result |>
  group_by(id) |>
  # the n() function counts the number of rows in each group
  summarise(n = n()) |>
  # isolate the groups that have more than 1 row
  filter(n > 1)
# we get a non-empty dataset, which means that there are duplicate rows in the result
sanity_check_1

# check that all students are in the result
# -> they're not, because student 2 is missing
sanity_check_2 <- all(students$id %in% result$id)
sanity_check_2
```

## Fixing problems with joins / the difference between left- and inner-joins. 

Fixing problems is really specific to the task at hand, and the problems you encountered in your diagnostics. 

Here, we have two problems:

- Student 2 didn't take the test, thus has no grade -> we will discuss this later
- Student 3 took the test twice, thus has two grades -> we will aggregate his grades by taking the mean of his two test scores.
- Student 4 took the test, but is not in the students dataset -> we want to ignore this student

### Removing duplicates

```{r}
#| label: fixing-joins

# first, let's fix the problem of student 3 having two grades by taking the mean of his two test scores
tests_fixed <- tests |>
  group_by(id) |>
  summarise(test_score = mean(test_score))
```

### Handling missing data

When doing a join, it is common to distinguish the "left" table and the "right" table. When you write `inner_join(students, tests, by = "id")`, the students dataset is the left table, and the tests dataset is the right table. The type of join you choose (e.g., inner join, left join, right join) determines how you handle missing data.

- `inner_join()`: only keeps rows that have a match in both datasets. In our example, this means that we will only keep students who have a test score, and we will only keep test scores that have a corresponding student. This is the most common type of join.
- `left_join()`: keeps all rows from the left dataset, and only the matching rows from the right dataset. In our example, this means that we will keep all students, even if they don't have a test score, but we will only keep test scores that have a corresponding student.
- There are two less common types of joins: `right_join()` and `full_join()`. `right_join()` is the opposite of `left_join()`: it always  We will not cover them in this lecture, but you can read about them in the documentation.
- There is one last, useful type of join: `anti_join()`. This keeps only the rows that do not have a match in the other dataset. In our example, this means that we will keep only students who don't have a test score, and only test scores that don't have a corresponding student. It is nice for diagnosing problems with joins, because it allows us to see which rows are causing the problems.

In practice, you will really balance between `inner_join()` and `left_join()`. In our use case: 

- Say you want to compute mean scores by gender: you don't care about students who don't have a test score, so you can use `inner_join()`.
- Say you want to do a mailing list where you want to send an email to all students, with the grade for those who have one, and "you don't have a grade" for those who don't have a test score, then you can use `left_join()`.

```{r}
# keep only students who are in both the left and right datasets
inner_join(students, tests_fixed, by = "id")

# keep all students in the left dataset, and only the matching rows in the right dataset
left_join(students, tests_fixed, by = "id")

# keep all students (not very useful, but just for demonstration purposes)
full_join(students, tests_fixed, by = "id")

# useful for debugging: keep all the students who were not matched in the right dataset
anti_join(students, tests_fixed, by = "id")
```

## Subtelties of joins

### Joining on variables with different names

Sometimes, the variable that we want to join on has different names in the two datasets. For example, in the students dataset, the variable that identifies the student is called "id", while in the tests dataset, it is called "student_id" (because in the tests dataset, the id column refers to the test itself). There are a few ways to handle that. 

```{r}
#| label: different-names

tests <- tibble(
  student_id = c(1, 1, 2, 2, 3, 3),
  test_score = c(90, 85, 95, 80, 10, 20),
  id = c(1, 2, 1, 2, 1, 2)
)

# approach 1: specify the by argument with the different variable names
# left is for the left dataset, right is for the right dataset
inner_join(students, tests, by = c("id" = "student_id"))

# approach 2: rename the variable in one of the datasets so that they have the same name, and then join
# (that's my favorite approach, because it makes the code more readable)
inner_join(
  students,
  tests |> rename(test_id = id, id = student_id),
  by = "id"
)
```

### Joining on multiple variables

Sometimes, you don't have a unique identifier that you can join on, but you have multiple variables that together identify the rows. For example, in the students dataset, we have the name and age of the student, while in the tests dataset, we also have the name and age of the student. We can use both the name and age variables to perform the join. This is not a good practice, because it is prone to errors (e.g., if there are typos in the names, or if there are multiple students with the same name and age). But sometimes, you don't have a choice. 

```{r}
#| label: multiple-variables

students <- tibble(
  id = c(1, 2, 3),
  name = c("Alice", "Bob", "Charlie"),
  age = c(20, 21, 22),
  email = c("alice@example.com", "bob@examoke.com", "charlie@me.com")
)

# see how our tests data has two Alices:
# young Alice (also in the students dataset)
# and old Alice (not in the students dataset)
tests <- tibble(
  name = c("Alice", "Bob", "Charlie", "Alice"),
  age = c(20, 21, 22, 23),
  test_score = c(90, 85, 95, 10)
)

# when joining, we need to specify both the name and age variables in the by argument
# notice how old Alice gets dropped from the result, because there is no matching row in the students dataset
inner_join(students, tests, by = c("name", "age"))
```

# Pivoting

There are two formats: 

- The long format: each row is an observation, and each column is a variable. This is the format that we typically use for data analysis and modeling.
- The wide format: each row contains several observations, encoded as different columns.

## Pivoting from wide to long format

```{r}
#| label: wide-to-long

wide <- tibble(
  id = c(1, 2, 3),
  name = c("Alice", "Bob", "Charlie"),
  test_1 = c(90, 95, 10),
  test_2 = c(85, 80, 20)
)

longer <- wide |>
  rename(first_name = name) |>
  pivot_longer(
    # the columns you want to pivot are the test_1 and test_2 columns
    cols = c(test_1, test_2),
    # these arguments are optional, but they make the output more readable
    names_to = "test",
    values_to = "score"
  )

longer

# why would we want to pivot?
# because, say, we want to compute mean scores

longer |>
  group_by(id) |>
  summarise(mean_score = mean(score))
```

## Pivoting from long to wide format


```{r}
#| label: long-to-wide

long <- tibble(
  id = c(1, 1, 2, 2, 3, 3),
  name = c("Alice", "Alice", "Bob", "Bob", "Charlie", "Charlie"),
  test = c(1, 2, 1, 2, 1, 2),
  score = c(90, 85, 95, 80, 10, 20)
)

long |>
  pivot_wider(
    names_from = test,
    values_from = score,
    # this is optional, but it makes the output more readable
    names_prefix = "test_"
  )

# why would we want to pivot?
# say because we want both the mean score by student and the details of each test score in the same dataset
long |>
  group_by(id) |>
  mutate(
    gpa = mean(score)
  ) |>
  pivot_wider(
    names_from = test,
    values_from = score,
    names_prefix = "test_"
  )
```