---
title: "Lecture 12: Advanced data manipulation: joins, pivoting"
---

```{r}
#| label: setup

suppressPackageStartupMessages(library(tidyverse))
```

# Joins

Our problem: we have two datasets; one with information about students (e.g., name, age, grade), and another with information about their test scores (e.g., name, test score). We want to combine these two datasets into one dataset that contains all the information about the students and their test scores.

## Example 1: a one-to-one join

```{r}
#| label: student-data

students <- tibble(
  id = c(1, 2, 3),
  name = c("Alice", "Bob", "Charlie"),
  age = c(20, 21, 22)
)
```

## Example 2: a one-to-many join

```{r}
#| label: tests-data-2

tests <- tibble(
  id = c(1, 1, 2, 2, 3, 3),
  test_score = c(90, 85, 95, 80, 10, 20)
)
```

## Diagnosing problems with joins

```{r}
#| label: tests-data-3

tests <- tibble(
  id = c(1, 3, 3, 4),
  test_score = c(90, 85, 95, 10)
)
```

## Subtelties of joins

### Joining on variables with different names

```{r}
#| label: different-names

tests <- tibble(
  student_id = c(1, 1, 2, 2, 3, 3),
  test_score = c(90, 85, 95, 80, 10, 20)
)
```

### Joining on multiple variables

```{r}
#| label: multiple-variables

students <- tibble(
  id = c(1, 2, 3),
  name = c("Alice", "Bob", "Charlie"),
  age = c(20, 21, 22)
)

tests <- tibble(
  name = c("Alice", "Bob", "Charlie"),
  age = c(20, 21, 22),
  test_score = c(90, 85, 95)
)
```

# Pivoting

There are two formats: 

- The long format: each row is an observation, and each column is a variable. This is the format that we typically use for data analysis and modeling.
- The wide format: each row contains several observations, encoded as different columns.

## Pivoting from wide to long format

```{r}
#| label: wide-to-long

wide <- tibble(
  id = c(1, 2, 3),
  name = c("Alice", "Bob", "Charlie"),
  test_1 = c(90, 95, 10),
  test_2 = c(85, 80, 20)
)
```

## Pivoting from long to wide format


```{r}
#| label: long-to-wide

long <- tibble(
  id = c(1, 1, 2, 2, 3, 3),
  name = c("Alice", "Alice", "Bob", "Bob", "Charlie", "Charlie"),
  test = c(1, 2, 1, 2, 1, 2),
  score = c(90, 85, 95, 80, 10, 20)
)
```