---
title: "Lecture 17: Missing data and measurement error"
format: revealjs
---

```{r}
#| label: setup
#| echo: false

suppressPackageStartupMessages(library(tidyverse))
library(modelsummary)
source("./lectures/helpers.R")

df <- read_csv("./data/lecture-7-gss.csv", show_col_types = FALSE) |>
  clean_gss()
```

## Data quality problems

- **Measurement error**: we record a value, but it's wrong
  - Example: thermometer error → we record 99°C instead of 100°C
- **Missing data**: we fail to record a value at all
  - Example: respondent refuses to report income → we have no value for income

Problems can be:

- **Random**: unsystematic — adds noise but not bias
  - Example: thermometer error that sometimes reads high and sometimes low
- **Systematic**: follows a pattern — biases estimates
  - Example: thermometer error that always reads 1°C too high

**Which is worse?** 

. . .

Systematic — but random is still a problem.

# Measurement error

## Thinking about measurement error

Statistics starts with the idea that every measurement contains some sampling/measurement error:

$$\tilde{y} = y + \varepsilon$$

- $y$ = true (unobserved) value
- $\tilde{y}$ = value we record
- $\varepsilon$ = measurement error

. . .

**Those two situations are analoguous**

- $y$ = true boiling point of water; $\tilde{y}$ = one measurement by an imperfect thermometer; $\varepsilon$ = thermometer error → **measurement** error
- $y$ = average class height; $\tilde{y}$ = one student's height; $\varepsilon$ = **sampling** error

## Why averaging works

```{r}
#| label: measurement-error
#| fig-height: 4

set.seed(42)
tibble(measurement = rnorm(1000, mean = 100, sd = 2)) |>
  ggplot(aes(x = measurement)) +
  geom_density(fill = "steelblue", alpha = 0.4) +
  geom_vline(
    xintercept = 100,
    linetype = "dashed",
    color = "red",
    linewidth = 1
  ) +
  annotate(
    "text",
    x = 100.3,
    y = 0.15,
    label = "True value (100°C)",
    hjust = 0,
    color = "red"
  ) +
  scale_x_continuous(limits = c(93, 107)) +
  labs(
    title = "1,000 measurements of the boiling point of water",
    x = "Measured temperature (°C)",
    y = "Density"
  ) +
  theme_minimal()
```

Because errors are random (some high, some low), averaging many $\tilde{y}_i$ cancels them out and recovers $y$. 

The law of large numbers helps quantify the uncertainty (p-values, confidence intervals, etc).

. . .

**This only works when errors are truly random.**

## Random measurement error

Two research questions:

1. At what temperature does water boil? → estimate $y = \tilde{y} + \epsilon$

. . .

| Source of error | Consequence | Fix |
|---|---|---|
| $y$ | Noisy $y$, but unbiased | More observations or Better instrument |

. . .

2. How does boiling point vary with salinity? → estimate $\beta$ in $y = \alpha + \beta x + \varepsilon$

. . .

| Source of error | Consequence | Fix |
|---|---|---|
| $y$ | Noisy $\hat{\beta}$, but unbiased | More observations or Better instrument |
| $x$ | $\hat{\beta}$ biased toward zero | Better instrument only |

The case with measurement error in $x$ is called **attenuation bias**.

## Attenuation bias: intuition

- $x = \tilde{x} + \varepsilon$
- Different values of $x$ may reflect two things: real variation in salinity *and* random noise
- Regression cannot separate them
- So the relationship appears weaker than it really is $\Rightarrow$
- The estimated slope $\hat{\beta}$ shrinks toward zero

. . .

**More data does not help.** Only measuring $x$ more accurately does.

## Systematic measurement error

$$
\tilde{y} = y + c + \varepsilon
$$

- $c$ = systematic error (constant bias)
- $\varepsilon$ = random error

. . .

**Example:** thermometers from one supplier all read 1°C too high: $\tilde{y} = y + 1 + \varepsilon$

- Every measurement is wrong in the same direction
- Averaging gives a more precise estimate of the *wrong* value
- No amount of data fixes this

## Social science examples: random or systematic?

1. **"On a scale of 1–10, how happy are you?"**
2. **"Have you ever paid a bribe?"**

. . . 

1. Random error: mood on the day of the survey adds noise, but averages out across respondents
2. Systematic error: respondents systematically underreport socially undesirable behavior → we underestimate true prevalence

## Solutions to systematic measurement error

1. **Use better tools**: survey designs that reduce social desirability bias

2. **Speculate about how bias affects your results**

   - Finding: government trust = 0.5 (non-Trump supporters) vs 0.3 (Trump supporters)
   - Problem: non-Trump supporters may give more socially desirable answers

. . .

Interpretation: the *true* gap is likely **larger** than 0.2

# Missing data

## Missing data as extreme measurement error

- Error so large the value cannot be recorded at all
- The same taxonomy applies:

. . .

**Random**

- Missingness unrelated to any variable
- Sample is smaller and noisier, but not biased
- Example: survey respondents randomly skip some questions (e.g., due to fatigue)

. . .

**Systematic**

- Missingness related to the unobserved value
- Estimates are biased — more data does not help
- Example: survey respondents with high income refuse to report it → we underestimate average income

## Solutions to missing data

1. **Complete-case analysis**: drop observations with any missing value
   - Simple, but reduces $n$ and may bias estimates (if missingness is systematic)

2. **Missingness as a category**: create a binary indicator (= 1 when missing)
   - Diagnostic: if the indicator correlates with the outcome, data are likely MNAR

3. **Imputation**: fill in missing values with predicted estimates
   - *Mean imputation*: easy, but doesn't fix systematic missingness
   - *Regression imputation*: predict the missing variable from other observed variables

## Why imputation is useful

An observation has 9 of 10 variables observed. Throwing it away wastes 9 good values.

Impute the 10th from the other 9 → use the full observation.

. . .

**The caveat:** imputation is only as good as the model used. When missingness is systematic, regression imputation inherits the bias.

# In practice

## Two rules

We rarely know whether problems are random or systematic. Standard practice:

1. Show results **with and without controls**
2. Keep the **sample size constant** across all specifications

. . .

The second point is often overlooked. If one model uses 500 observations and another uses 800, changes in coefficients could reflect the *sample*, not the controls.

## A bad example

```{r}
#| label: bad-example
#| output-location: slide

modelsummary(
  list(
    "Education only" = lm(income ~ edu, data = df),
    "Education + Race + Age" = lm(income ~ edu + race + age, data = df)
  ),
  stars = TRUE,
  gof_map = c("nobs", "r.squared"),
  title = "Bad: sample sizes differ across models"
)
```

## Good example 1: fix the sample first

```{r}
#| label: good-example-1
#| output-location: slide

df_complete <- df |> drop_na(income, edu, race, age)

modelsummary(
  list(
    "Education only" = lm(income ~ edu, data = df_complete),
    "Education + Race + Age" = lm(income ~ edu + race + age, data = df_complete)
  ),
  stars = TRUE,
  gof_map = c("nobs", "r.squared"),
  title = "Better: same sample across all models"
)
```

## Good example 2: also show the full sample

```{r}
#| label: good-example-2
#| output-location: slide

modelsummary(
  list(
    "Full sample" = lm(income ~ edu, data = df),
    "Complete cases" = lm(income ~ edu, data = df_complete),
    "Complete cases + Controls" = lm(
      income ~ edu + race + age,
      data = df_complete
    )
  ),
  stars = TRUE,
  gof_map = c("nobs", "r.squared"),
  title = "Best: full sample, complete cases, and controls"
)
```

## Reading the three-column table

- **Col 1 → Col 2**: effect of restricting to complete cases (sample change only)
- **Col 2 → Col 3**: effect of adding controls, holding the sample fixed

. . .

This decomposition is what a careful reader needs to trust your results.

# To remember

1. Two types of data quality problems: measurement error and missing data
2. Both can be random (adds noise) or systematic (biases estimates)
3. Random problems can be fixed with more data (except attenuation bias)
4. Systematic problems usually require better quality data or careful interpretation
5. Show results with and without controls, and keep the sample size constant across all specifications

# An aside: linear regression vs. logistic regression

## A common misconception

- Linear regression: $y = \alpha + \beta x + \varepsilon$
- Logistic regression: $y = \frac{1}{1+\exp[-(\alpha + \beta x + \varepsilon)]}$

> My variable of interest is binary, so I have to use logistic regression.

. . .

**Not necessarily.** Linear regression can be used with binary outcomes, and often performs just as well as logistic regression.

## Advantages of linear regression with binary outcomes

1. Simpler to interpret; linear regression is just conditional means. Example: 
  - Approves of Putin = 1; disapproves = 0
  - Watches TV = 1; doesn't watch TV = 0
  - $\text{approves}_i = 0.4 + 0.1 \times \text{TV}_i$
  - Interpretation: 
    - **mean approval** for/probability of approving of Putin is 40% for non-TV watchers
    - **mean approval** for/probability of approving of Putin is 50% for TV watchers
2. In practice, usually gives similar results to logistic regression

## You don't need the main advantage of logistic regression

1. Logistic regression constrains predicted probabilities to [0, 1]
  - Approves of Putin = 1; disapproves = 0
  - $\text{approves}_i = -0.3 + 0.01 \times \text{age}_i$
  - Interpretation: the probability of approving Putin increases by 1 percentage point for every year of age
  - Problem: respondents of age 18 have a predicted probability of -0.12, which is impossible
  - Logistic regression fixes this: $\frac{1}{1 + \exp(x)} \in (0,1)$
2. If you don't need well-defined predictions, then linear regression is fine. 
  - Typically, a statement like "older people are more likely to approve of Putin" is sufficient.
