---
title: "Lecture 13: Causality"
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(modelsummary)
source("./lectures/helpers.R")

df <- read_csv("./data/lecture-7-gss.csv") |> clean_gss()
df_resume <- read_csv("./data/lecture-13-resume.csv")
```

## Empirical Motivation: Why Do Coefficients Change?

Consider a simple question: do Black and White Americans earn different incomes? We can answer this with a regression using the General Social Survey (GSS).

Recall that OLS with a binary regressor estimates a difference in group means: regressing income on a Black/White indicator simply recovers the average income gap between the two groups. Let's run that regression, and then add education as a control.

```{r}
#| label: models-race-income

df_bw <- df |> filter(race %in% c("Black", "White"))

modelsummary(
  list(
    "Race only" = lm(income ~ race, data = df_bw),
    "+ Education" = lm(income ~ race + edu, data = df_bw)
  ),
  stars = TRUE,
  gof_map = c("nobs", "r.squared")
)
```

Model 1 gives the raw Black–White income gap (the difference in group means). Model 2 adds education — and the coefficient on race shrinks substantially. Why?

To understand what is happening, let's build intuition with simulated data where we know the true data-generating process.

### Illustration with simulated data

We consider two worlds. In both, income is generated by the same rule: it depends on race and education. The only difference is whether race and education are correlated with each other.

```{r}
#| label: sim-data

set.seed(42)
n <- 1000

# World 1: race and education are independent
sim1 <- tibble(
  black = rbinom(n, 1, 0.5),
  edu = rnorm(n, mean = 12, sd = 3), # same distribution for both groups
  income = 20000 - 3000 * black + 1500 * edu + rnorm(n, sd = 5000)
)

# World 2: Black respondents have 2 fewer years of education on average
sim2 <- tibble(
  black = rbinom(n, 1, 0.5),
  edu = rnorm(n, mean = 12 - 2 * black, sd = 3), # correlated with race
  income = 20000 - 3000 * black + 1500 * edu + rnorm(n, sd = 5000)
)
```

```{r}
#| label: sim-regressions

mods <- list(
  lm(income ~ black, data = sim1),
  lm(income ~ black + edu, data = sim1),
  lm(income ~ black, data = sim2),
  lm(income ~ black + edu, data = sim2)
)

modelsummary(
  mods,
  stars = TRUE,
  gof_map = c("nobs", "r.squared")
) |>
  tinytable::group_tt(
    j = list(
      "Uncorrelated" = 2:3,
      "Correlated" = 4:5
    )
  )
```

In World 1 (left panel), adding education leaves the race coefficient essentially unchanged: because race and education are independent, omitting education does not bias the race estimate.

In World 2 (right panel), the race coefficient shrinks substantially when education is added. Because Black respondents have less education on average, the race-only model conflates the direct effect of race with the education penalty. Adding education separates the two.

**This is exactly the pattern we see in the GSS data.** Black respondents have, on average, fewer years of education than White respondents. The two variables are correlated, so adding education shifts the race coefficient. Whether that shift reflects "explaining" or "absorbing" the gap is a causal question — one that regression alone cannot answer. At any rate, we can say that **education confounds the relationship between race and income.**

**To remember.** When we add a control in an OLS regression, the coefficient changes only if the control is correlated with the main regressor. What that change reflects requires causal thinking — not just more controls.

## Counterfactuals and the Fundamental Problem of Causal Inference

Causality is about understanding the effect of a **treatment** on an **outcome**. Suppose we try to understand the impact of Advil (our treatment) on body temperature (the outcome). Denote $Y_i$ the temperature of subject $i$, and $T_i \in \{0,1\}$ whether subject $i$ took the treatment ($T_i = 1$) or not ($T_i = 0$). We can think of $Y_i(T_i)$: the subject's temperature given her treatment status. Let's say:

$$
Y_i(1) = 37^\circ C \quad \text{and} \quad Y_i(0) = 38^\circ C
$$

The **treatment effect** of Advil on her body temperature is $Y_i(1) - Y_i(0) = -1^\circ C$.

The problem is that we can never observe both $Y_i(1)$ and $Y_i(0)$ for the same subject. We can only observe one of them, depending on whether the subject took Advil or not. The outcome we do not observe is called the **counterfactual** outcome. This is the **fundamental problem of causal inference**: we cannot observe the counterfactual outcome.

You may wonder why we do not observe the counterfactual outcome. Specifically, at 3pm, you felt feverish and thus took an Advil. At 5pm, you measured your temperature again and noticed it had dropped. You thus concluded that Advil made your temperature drop. In reality, you made an assumption here: **you assumed that the only thing that changed between the two times you measured your temperature was taking the Advil.** Based on this assumption, you concluded that the drop in temperature was due to the Advil. But what if something else changed between the two times you measured your temperature? For example, what if you also took a nap, and that is what made your temperature drop? In that case, the drop in temperature could possibly not be due to the Advil, but rather to the nap. Here **the proper counterfactual is the following: what would your temperature have been at 5pm had you not taken the Advil?** We cannot observe this counterfactual, and thus we cannot be sure that the drop in temperature was due to the Advil.

**To remember.** The individual treatment effect is $\Delta_i = Y_i(1) - Y_i(0)$. We cannot estimate it because we observe either $Y_i(1)$ or $Y_i(0)$, but never both. The outcome we do not observe is the counterfactual. The fundamental problem of causal inference is that we cannot observe the counterfactual.

## The Average Treatment Effect and Why Randomization Solves the Problem

### The intuition

Instead of trying to estimate the individual treatment effect $\Delta_i = Y_i(1) - Y_i(0)$, we can try to estimate the **average treatment effect (ATE)** across a population. To do so, we run a **randomized controlled trial (RCT)**: we randomly assign subjects to either a treatment group ($T_i = 1$) or a control group ($T_i = 0$).

Why does randomization help? Return to the Advil example. Suppose we compare two individuals, $i$ and $j$. $i$ took Advil and $j$ did not. We find $Y_i = 37^\circ C$ and $Y_j = 36.3^\circ C$. We might be tempted to conclude that Advil *increases* temperature. But this comparison is invalid for two reasons:

1. **Selection**: $i$ took Advil *because* she was feverish. She selected into treatment. Her $Y_i(0)$ — what her temperature would have been without Advil — is $38^\circ C$, not $36.3^\circ C$.
2. **Confounding**: $j$ may be younger and naturally have a lower resting temperature. The difference between $i$ and $j$ could be due to age, not Advil.

Randomizing assignment **across a large number of individuals** solves both problems at once. Because we assign who takes Advil by a coin flip:

- There is no selection: treatment is not chosen by subjects, so those assigned to treatment are not systematically different from those assigned to control.
- There is no confounding: random assignment ensures that, on average, the treatment and control groups are identical in *all* characteristics — observed or unobserved. There are as many young/old and sick/healthy people in the treatment group as in the control group.

This is the key property of randomization: **treatment becomes uncorrelated with everything else** — including potential outcomes. It is as if we created two groups that are otherwise identical, and the only difference between them is whether they received the treatment.

### The math

We want to estimate the ATE:

$$
ATE = \mathbb{E}[Y_i(1) - Y_i(0)]
$$

We never observe $Y_i(1)$ and $Y_i(0)$ for the same individual simultaneously. But we can try to estimate $\mathbb{E}[Y_i(1)]$ and $\mathbb{E}[Y_i(0)]$ separately, using the treated and control groups respectively.

The challenge is that the treated group is not the whole population: it is the subgroup with $T_i = 1$. So what we observe is $\mathbb{E}[Y_i(1) \mid T_i = 1]$, not $\mathbb{E}[Y_i(1)]$. In general, these two quantities are different — the treated group is a selected sample. In the Advil example, the treated group consists mainly of sick people, so $\mathbb{E}[Y_i(1) \mid T_i = 1] > \mathbb{E}[Y_i(1)]$.

With **random assignment**, however, treatment $T_i$ is independent of potential outcomes $(Y_i(1), Y_i(0))$:

$$
T_i \perp (Y_i(1), Y_i(0))
$$

This is precisely the mathematical statement that treatment is uncorrelated with everything. And it has a decisive consequence: conditioning on $T_i$ tells us nothing about $Y_i(1)$ or $Y_i(0)$, so:

$$
\mathbb{E}[Y_i(1) \mid T_i = 1] = \mathbb{E}[Y_i(1)]
\qquad \text{and} \qquad
\mathbb{E}[Y_i(0) \mid T_i = 0] = \mathbb{E}[Y_i(0)]
$$

The treated group is a representative sample of the population for $Y_i(1)$, and the control group is a representative sample for $Y_i(0)$. *It is only because of this independence that* we can write:

$$
\mathbb{E}[Y_i(1)] - \mathbb{E}[Y_i(0)] = \mathbb{E}[Y_i(1) \mid T_i = 1] - \mathbb{E}[Y_i(0) \mid T_i = 0]
$$

### From expectations to sample averages

By the **Law of Large Numbers**, with enough observations, sample averages converge to population expectations:

$$
\bar{Y}_1 = \frac{1}{n_1}\sum_{i:\, T_i=1} Y_i \approx \mathbb{E}[Y_i(1) \mid T_i = 1] = \mathbb{E}[Y_i(1)]
$$

Putting it all together:

$$
\widehat{ATE} = \bar{Y}_1 - \bar{Y}_0 \approx \mathbb{E}[Y_i(1)] - \mathbb{E}[Y_i(0)] = ATE
$$

**The difference in sample means is a valid estimator of the ATE — only because of random assignment.**

In our Advil example: suppose the experiment yields $\bar{Y}_1 = 37^\circ C$ in the treatment group and $\bar{Y}_0 = 37.5^\circ C$ in the control group. The $37.5^\circ C$ reflects the fact that the control group contains both healthy people (temperature $\approx 37^\circ C$) and feverish people (temperature $\approx 38^\circ C$). The treatment group, being equally mixed by random assignment, would also have averaged $37.5^\circ C$ without the drug — but Advil brought feverish subjects down to $37^\circ C$. The estimated ATE is $37 - 37.5 = -0.5^\circ C$.

**To remember.** Randomization makes treatment uncorrelated with potential outcomes: $T_i \perp (Y_i(1), Y_i(0))$. This is what allows us to use the treated group to estimate $\mathbb{E}[Y_i(1)]$ and the control group to estimate $\mathbb{E}[Y_i(0)]$. The difference in sample means then estimates the ATE — but only under random assignment.

---

## Heterogeneous Treatment Effects

The ATE averages the treatment effect across the entire population. But treatment effects often vary across subgroups — we say they are **heterogeneous**.

Return to the Advil example. Suppose we estimate the ATE separately for healthy and sick people:

- For **healthy** people: $\bar{Y}_1 = 37^\circ C$ and $\bar{Y}_0 = 37^\circ C$, so $ATE = 0$.
- For **sick** people: $\bar{Y}_1 = 37^\circ C$ and $\bar{Y}_0 = 38^\circ C$, so $ATE = -1^\circ C$.

Advil works for sick people, not for healthy people. The overall ATE pools these two groups together and gives a number somewhere in between — which is informative for average policy, but masks who actually benefits.

Knowing *who* gets treated matters. If a drug is only assigned to healthy people, the overall ATE will understate the effect for those who actually need it.

**To remember.** Treatment effects can vary across subgroups. We can estimate subgroup ATEs by repeating the randomization logic within each group, as long as we measure the relevant characteristics before treatment assignment.

---

## Application: The Bertrand Experiment

### Context

Bertrand & Mullainathan (2004) ran a field experiment to measure racial discrimination in the labor market. They sent fictitious résumés to real job postings in Boston and Chicago. The résumés were otherwise identical, but the **name** on the résumé was randomly assigned to signal race: stereotypically Black names (Lakisha, Jamal) or White names (Emily, Greg). The outcome was whether the employer called back.

This is a genuine experiment: the name — and thus the race signal — was randomly assigned. So:

$$
T_i \perp (Y_i(1), Y_i(0))
$$

The difference in callback rates between groups is therefore a valid causal estimate of the effect of race on employer callbacks.

### Analysis

Recall that OLS with a binary regressor estimates a difference in group means. Here `race` is binary (black/white), so regressing `call` on `race` directly recovers $\bar{Y}_{\text{white}} - \bar{Y}_{\text{black}}$ — the estimated ATE.

```{r}
#| label: resume-ols

modelsummary(
  lm(call ~ race, data = df_resume),
  stars = TRUE,
  gof_map = c("nobs", "r.squared")
)
```

The intercept is the callback rate for Black-named résumés ($\bar{Y}_{\text{black}}$); the coefficient on `racewhite` is the gap — the estimated ATE. White-named résumés receive substantially more callbacks than Black-named résumés. The gap is large and highly statistically significant. Because the name was **randomly assigned**, this is a causal estimate: it is the effect of the race signal, not of any other characteristic.

### Connecting back to the GSS

In the GSS analysis, we saw that the Black–White income gap shrinks once we control for education. But we could not determine whether this reflected causation or confounding. Here, we *can* make a causal claim: the résumés were identical except for the name, so the gap in callbacks can only be due to the name — i.e., to the perceived race of the applicant.

This illustrates the power of random assignment. Without it (as in the GSS), we are always unsure whether our estimate reflects causation or the fact that treated and untreated individuals differ in unobserved ways.

**To remember.** The Bertrand experiment estimates the causal effect of perceived race on employer callbacks. Random assignment of names guarantees that $\bar{Y}_1 - \bar{Y}_0$ is a valid ATE estimate. The result — a large, significant callback gap — provides direct evidence of racial discrimination in hiring.
