---
title: "Lecture 8-9: advanced linear regression, tables and plots"
---

```{r}
#| label: setup

suppressPackageStartupMessages(library(tidyverse))
library(modelsummary)
library(broom)
library(marginaleffects)
# library(knitr) # we will use only one function from this package, so we can load it separately
df <- read_csv("./data/lecture-7-gss.csv", show_col_types = FALSE)
```

```{r}
#| label: data-cleaning

df <- df |>
  mutate(
    # trun sex into a binary variable that = 1 for female,
    # 0 for male, NA for everything else
    female = case_match(
      sex,
      "FEMALE" ~ 1,
      "MALE" ~ 0,
      .default = NA
    ),
    # remove negative values from height, and convert from inches to cm
    height = ifelse(rheight < 0, NA, rheight),
    height = height * 2.54,
    # turn income bins into a numeric variable by taking the midpoint of each bin
    # for the top bin, we assign it a value of 25000
    # for the bottom bin, we assign it a value of 1000
    income = case_match(
      rincome,
      "$25000 OR MORE" ~ 25000,
      "$20000 - 24999" ~ 22500,
      "$15000 - 19999" ~ 17500,
      "$10000 - 14999" ~ 12500,
      "$8000 TO 9999" ~ 9000,
      "$7000 TO 7999" ~ 7500,
      "$6000 TO 6999" ~ 6500,
      "$5000 TO 5999" ~ 5500,
      "$4000 TO 4999" ~ 4500,
      "$3000 TO 3999" ~ 3500,
      "$1000 TO 2999" ~ 2000,
      "LT $1000" ~ 1000,
      .default = NA
    ),
    # race: we keep only the three meaningful categories,
    # and set the rest (missing race, ...) to NA
    race = ifelse(race %in% c("Black", "White", "Other"), race, NA),
    # turn education categories into a numeric variables
    # (i.e., years of education)
    edu = case_match(
      educ,
      "No formal schooling" ~ 0,
      "1st grade" ~ 1,
      "2nd grade" ~ 2,
      "3rd grade" ~ 3,
      "4th grade" ~ 4,
      "5th grade" ~ 5,
      "6th grade" ~ 6,
      "7th grade" ~ 7,
      "8th grade" ~ 8,
      "9th grade" ~ 9,
      "10th grade" ~ 10,
      "11th grade" ~ 11,
      "12th grade" ~ 12,
      "1 year of college" ~ 13,
      "2 years of college" ~ 14,
      "3 years of college" ~ 15,
      "4 years of college" ~ 16,
      "5 years of college" ~ 17,
      "6 years of college" ~ 18,
      "7 years of college" ~ 19,
      "8 or more years of college" ~ 20,
      .default = NA
    ),
    # age: we set the top bin to 89, and convert to numeric
    age = ifelse(age == "89 or older", 89, age),
    # converting to numeric will turn "missing" into an NA automatically
    # R will give us a warning when doing this
    # we can ignore the warning, because we actually want the "missing" values to be turned into NAs
    age = as.numeric(age)
  ) |>
  select(age, female, height, income, edu, race)
```

## Model fit: the R-squared

The $R^2$ is a common measure of model fit for linear regression. It is the proportion of variance in the outcome variable that is explained by the model. The $R^2$ can be calculated as:

$$
R^2 = 1 - \frac{SS_{res}}{SS_{tot}}
$$

Where $SS_{res}$ is the sum of squared residuals (so, the sum of $\epsilon_i^2$'s; i.e., the difference between the observed and predicted values) and $SS_{tot}$ is the total sum of squares (the variance of the outcome variable). The $R^2$ ranges from 0 to 1, with higher values indicating a better fit of the model to the data.

Adding variables to the model will always (weakly) increase the $R^2$, even if those variables are not actually related to the outcome variable. In other words, say you estimate model 1, and call its R-squared $R^2_1$. Add variables to model 1, and call the new model model 2, and its R-squared $R^2_2$. We have

$$
R^2_2 \geq R^2_1
$$

This is because if a variable doesn't improve the fit at all, we assign it a coefficient of 0, and the $R^2$ won't change. Example: 

```{r}
#| label: r-squared

df_mini <- df |> select(height, income, female) |> na.omit()

model_1 <- lm(income ~ height, data = df_mini)
model_2 <- lm(income ~ height + female, data = df_mini)

# access the R-squared values from the model summaries
summary(model_1)$r.squared # r squared for model 1
summary(model_2)$r.squared # r squared for model 2; it is higher than model 1
```

To adjust for that, we can use the adjusted $R^2$, which penalizes the addition of variables to the model. Most people don't care much about the adjusted $R^2$.

## Making tables

The function `kable` from the `knitr` package is a simple way to create tables in R. It takes a data frame as input and returns a formatted table. Typically, `kable` is the only function you'll be using from the `knitr` package, so you can load it directly without loading the entire package, using notation `knitr::kable()`.

```{r}
#| label: basic-table
#| tbl-cap: Basic table

# you can use kable to create a simple table from a data frame
# you can use block options to add a caption to the table.
# Specifically, the option tbl-cap adds a caption to the table
knitr::kable(head(df))
```

The function `modelsummary` from the `modelsummary` package is a powerful way to create regression tables in R. It takes a list of models as input and returns a formatted table with the coefficients, standard errors, and other statistics for each model. You can customize the table using various options, such as adding stars to indicate statistical significance.

```{r}
#| label: regression-tables
#| tbl-cap: Our first regression table

modelsummary(
  list(model_1, model_2),
  stars = TRUE
)
```

```{r}
#| label: customize-regression-tables
#| tbl-cap: A better regression table

modelsummary(
  list(
    # you can give custom names to the models by using a named list
    # notice how I use backticks to allow for spaces in the model names
    `Baseline model` = model_1,
    `Larger model` = model_2
  ),
  stars = TRUE,
  # you can tweak the statistics that are displayed in the table using the gof_map argument
  # here, we only display the number of observations (nobs) and the R-squared (r.squared)
  gof_map = c("nobs", "r.squared")
)
```

## Categorical variables

Let's try and get a sense of how race affects income. Notice that race is a **categorical variable**: there are three "categories" (White, Black, Other). Regression only uses numeric variables, so we need to convert the race variabble into a numeric variable. To do that, we will expand the idea of a binary variable that we used for the female variable. 

### Binary variables, take 2

A binary variable, also called a dummy variable, is a variable that takes the value of 0 or 1. Last time, we ran the following regression: 

$$
y_i = \alpha_1 + \beta_1 \text{female}_i + \epsilon_i
$$

We chose to assign the value 1 to female and 0 to male. This means that

- $\alpha_1$ is the average income for males
- $\beta_1$ is the difference in average income between females and males
- $\alpha_1 + \beta_1$ is the average income for females

We could have done the opposite, and assigned the value 1 to male and 0 to female. In that case, we would have had

$$
y_i = \alpha_2 + \beta_2 \text{male}_i + \epsilon_i
$$

In this case, we would have had: 

- $\alpha_2$ is the average income for females
- $\beta_2$ is the difference in average income between males and females
- $\alpha_2 + \beta_2$ is the average income for males

Notice that the two models are equivalent. In other words, we get the same predicted values for each observation, and the same $R^2$ value. The only thing that changes is the interpretation of the coefficients. Specifically, we have: 

- $\alpha_1 = \alpha_2 + \beta_2$: the average income for males
- $\alpha_2 = \alpha_1 + \beta_1$: the average income for females

What we really did, really was to change the **reference category** of our binary variable. In the first model, the reference category is male, because we assigned it the value 0. In other words, we get the effect of being female **relative to the reference** of being male. 

## More than two categories

When we have a categorical variable with more than two categories, we can use the same idea to create multiple dummy variables. We can turn our race variable into two dummy variables: one for Black and one for Other. The regression model will look like this:

$$
y_i = \alpha + \beta_1 \text{Black}_i + \beta_2 \text{Other}_i + \epsilon_i
$$

Why didn't we create a dummy variable for White? Because we need to choose a reference category, and we will choose White as the reference category. In this model, we have: 

- $\alpha$ is the average income for White people
- $\beta_1$ is the difference in average income between Black and White people
- $\beta_2$ is the difference in average income between Other and White people
- $\alpha + \beta_1$ is the average income for Black
- $\alpha + \beta_2$ is the average income for Other

If we had added a dummy variable for White, we would have run into a problem. Look at this wrong model: 

$$
y_i = \alpha + \beta_1 \text{Black}_i + \beta_2 \text{Other}_i + \beta_3 \text{White}_i + \epsilon_i
$$

The problem here is that it is unclear what the intercept $\alpha$ represents: there is some confusion between $\beta_3$ and $\alpha$. If we run a regression with this model, R will complain.

Just like with the binary variable, we could have chosen a different reference category. For example, we could have chosen Black as the reference category, and created dummy variables for White and Other. In that case, the model would look like this:

$$
y_i = \tilde{\alpha} + \tilde{\beta}_1 \text{White}_i + \tilde{\beta}_2 \text{Other}_i + \tilde{\epsilon}_i
$$

In this model, we have:

- $\tilde{\alpha}$ is the average income for Black people
- $\tilde{\beta}_1$ is the difference in average income between White and Black people
- $\tilde{\beta}_2$ is the difference in average income between Other and Black people
- $\tilde{\alpha} + \tilde{\beta}_1$ is the average income for White
- $\tilde{\alpha} + \tilde{\beta}_2$ is the average income for Other

Again, just like with the binary variable, the two models are equivalent. We get the same predicted values for each observation, and the same $R^2$ value. The only thing that changes is the interpretation of the coefficients. Specifically, we have:

- $\alpha = \tilde{\alpha} + \tilde{\beta}_1$: the average income for White people
- $\tilde{\alpha} = \alpha + \beta_1$: the average income for Black people
- $\tilde{\alpha} + \tilde{\beta}_2 = \alpha + \beta_2$: the average income for Other people

## Categorical variables in R

Now, let's do this in R. The function `lm` will automatically create dummy variables for us when we include a categorical variable in the model. By default, R will choose the reference category as the one that comes first alphabetically. In our case, that is Black. So, when we run the following regression, R will automatically create dummy variables for Other and White, and use Black as the reference category.

```{r}
#| label: categorical-variables

mod_race <- lm(income ~ race, data = df)
modelsummary(list(mod_race), stars = TRUE)
```

This is a little awkward to interpret, because there is a natural grouping between Other and Black. So we probably want to choose White as the reference category. It turns out R has a data type that is specifically designed for categorical variables, called a `factor`. A factor has `levels`, which are ordered. The first level is the reference category. We can convert our character variable `race` into a factor variable, and specify the reference category using the `fct_relevel` function.

```{r}
#| label: changing-reference-category

df <- df |>
  mutate(
    # convert race from a character variable to a factor variable
    race = as_factor(race)
  )
# check the levels of the factor variable
# the first level is the reference category
# it is the one that comes first alphabetically, which is "Black"
levels(df$race)

# change the reference category to "White"
df <- df |>
  mutate(
    race = fct_relevel(race, "White")
  )

# now, our regression model will use White as the reference category
mod_race <- lm(income ~ race, data = df)
modelsummary(list(mod_race), stars = TRUE)
```

We can visualize the coefficients of the model using a coefficient plot. A coefficient plot is a plot of the coefficients, with error bars representing the confidence intervals. 

```{r}
#| label: coef-plot

# the function broom::tidy() takes a model object and
# returns a data frame with the coefficients, standard errors, and confidence intervals
# for each term in the model. We can use this data frame to create a coefficient plot using ggplot2.
tidy(mod_race, conf.int = TRUE) |>
  # we remove the intercept, because it represents the average income for the reference category
  # this is very different from the other parameters, which represent
  # differences in average income between the reference category and the other categories
  filter(term != "(Intercept)") |>
  ggplot(aes(x = term, y = estimate)) +
  geom_point() +
  # let's add a dashed line at 0, to make it easier to see which coefficients are statistically significant
  geom_hline(yintercept = 0, linetype = "dashed") +
  # format the y-axis labels as monetary values
  scale_y_continuous(labels = scales::dollar) +
  # let's invert the axes, to take advantage of the horizontal space we have when writing papers
  coord_flip() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  labs(
    x = "Race",
    y = "Difference in average income (relative to White)"
  )
```

## Linear hypotheses

So, we have estimated the following model:

$$
y_i = \alpha + \beta_1 \text{Black}_i + \beta_2 \text{Other}_i + \epsilon_i
$$

So far, we know that Black people and Other race people have an income that is significantly lower than White people, because the coefficients $\beta_1$ and $\beta_2$ are negative and statistically significant. But there are a few things we don't know or would like to visualize. 

### Are Black people significantly poorer than Other race? 

The coefficient plot we just created only shows that the confidence intervals for Black and Other race parameters ($\beta_1$, $\beta_2$ respectively) overlap with each other. So there is no significant difference between the two groups. But we can also test this hypothesis directly. We want to know whether $\beta_1 \neq \beta_2$. The null hypothesis is that $\beta_1 = \beta_2$, which is equivalent to saying that $\beta_1 - \beta_2 = 0$. It turns out that you can't just add the p-values, or do the difference in confidence intervals. Instead, you need to test this new (linear) hypothesis separately. The `hypotheses` function from the `marginaleffects` package lets us do that.

```{r}
#| label: linear-hypothesis

hypotheses(mod_race, hypothesis = "raceOther - raceBlack = 0")

# the column Pr(>|z|) is the p-value. It is not significant, so we are confident that there is no significant difference between the two groups.
# the columns 2.5% and 97.5% give the confidence interval for the difference between the two groups.
```

### What is the average income for each group?

Our coefficient plot only shows the differences in average income between the reference category (White) and the other categories (Black, Other). It doesn't show the average income for each group. We can calculate the average income for each group using the coefficients from the model. The average income for White people is given by the intercept $\alpha$. We can get the confidence interval directly using `broom::tidy()`. The average income for Black people is given by $\alpha + \beta_1$. That's a linear hypothesis, so we can get the confidence interval using `marginaleffects::hypotheses()`. 

```{r}
#| label: linear-hypotheses

# get te plot data:
# a series of 1-row data frames, each containing the estimate and confidence interval for the average income of each group
# we combine them with bind_rows() to get a single data frame that we can use for plotting
pl <- bind_rows(
  # get the average income for White people, which is given by the intercept parameter
  tidy(mod_race, conf.int = TRUE) |>
    filter(term == "(Intercept)") |>
    # give it a more meaningful name, to make the plot easier to read
    mutate(term = "White"),
  # get the average income for Other race, which is given by the linear hypothesis (Intercept) + raceOther
  hypotheses(mod_race, hypothesis = "`(Intercept)` + raceOther = 0") |>
    mutate(term = "Other"),
  hypotheses(mod_race, hypothesis = "`(Intercept)` + raceBlack = 0") |>
    mutate(term = "Black")
)

# make the plot
ggplot(aes(x = term, y = estimate), data = pl) +
  geom_point() +
  scale_y_continuous(labels = scales::dollar) +
  coord_flip() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  labs(
    x = "Race",
    y = "Average income"
  )
```

## Interaction terms

Now, let's say we want to estimate returns to education. In other words, we want to know how much income increases for each additional year of education. We can estimate the following model:

$$
y_i = \alpha + \beta \text{edu}_i + \epsilon_i
$$

```{r}
#| label: returns-to-education

mod <- lm(income ~ edu, data = df)
summary(mod)
```

The model below shows that the returns to education are positive and statistically significant: one additional year of education increases income by about $`round(coef(mod)["edu"]`.

However, we might be interested in whether the returns to education are different for males and females. To answer that question, we can estimate two separate models; one for males and one for females. 

```{r}
#| label: returns-to-education-by-gender

models <- list(
  males = lm(income ~ edu, data = df |> filter(female == 0)),
  females = lm(income ~ edu, data = df |> filter(female == 1))
)

modelsummary(models, stars = TRUE)
```

The returns to education seem higher for females than for males: one additional year of education increases income by about $`round(coef(models[["males"]])["edu"])` for males, and $`round(coef(models[["females"]])["edu"])` for females. But is this difference statistically significant? We can eyeball it by looking at the confidence intervals for the two coefficients. 

```{r}
#| label: returns-to-education-confidence-intervals

models |>
  map_dfr(broom::tidy, conf.int = TRUE, .id = "model") |>
  filter(term == "edu") |>
  ggplot(aes(x = model, y = estimate)) +
  geom_point() +
  scale_y_continuous(labels = scales::dollar) +
  coord_flip() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  labs(
    x = "Model",
    y = "Returns to an additional year of education"
  )
```

We can also test the hypothesis that the returns to education are the same for males and females by running "two regressions in one". We can do that by including an **interaction term** between education and gender. An interaction term is a variable that is created by multiplying two variables together. It extends the idea of categorical variables: instead of just having a different intercept for each group, we also have a different slope for each group. The model looks like this:

$$
y_i = \alpha + \beta_1 \text{edu}_i + \beta_2 \text{female}_i + \beta_3 \text{edu}_i \times \text{female}_i + \epsilon_i
$$

Why does this work? Let's look at the model separately for males and females.

For males, we have $\text{female}_i = 0$, so the model becomes $y_i = \alpha + \beta_1 \text{edu}_i + \beta_2 \times 0 + \beta_3 \text{edu}_i \times 0 + \epsilon_i$. This simplifies to $y_i = \alpha + \beta_1 \text{edu}_i + \epsilon_i$; that is, a line with intercept $\alpha$ and slope $\beta_1$. So the returns to education for males are given by $\beta_1$. 

For females, we have $\text{female}_i = 1$, so the model becomes $y_i = \alpha + \beta_1 \text{edu}_i + \beta_2 \times 1 + \beta_3 \text{edu}_i \times 1 + \epsilon_i$. This simplifies to $y_i = (\alpha + \beta_2) + (\beta_1 + \beta_3) \text{edu}_i + \epsilon_i$; that is, a line with intercept $\alpha + \beta_2$ and slope $\beta_1 + \beta_3$. So the returns to education for females are given by $\beta_1 + \beta_3$. 

Notice furthermore that $\beta_3$ is the difference in returns to education between females and males. We call $\text{edu}_i \times \text{female}_i$ the **interaction** term, because it captures the interaction between education and gender. We call $\text{edu}_i$ and $\text{female}_i$ the **main effects**, because they capture the main effects of education and gender on income, respectively.

Now, let's see how to estimate models with interaction terms in R.

```{r}
#| label: interaction-terms

# the simple way: create a new variable that is the product of education and female
df <- df |> mutate(edu_female = edu * female)
lm(income ~ edu + female + edu_female, data = df) |> summary()

# the better way: use the : operator to create the interaction term automatically
lm(income ~ edu + female + edu:female, data = df) |> summary()

# the wizard's way: use the * operator to create
# the interaction term and the main effects automatically
# in other words, edu * female is expanded automatically
# into edu + female + edu:female
lm(income ~ edu * female, data = df) |> summary()
```

We notice that the coefficient on `edu:female` is positive and statistically significant, which means that the returns to education are higher for females than for males. 
