---
title: "Lecture 15: Cheating at regression"
---

```{r}
#| label: setup
#| echo: false

suppressPackageStartupMessages(library(tidyverse))
library(modelsummary)
```

## Empirical Motivation: Paul the Octopus

In the 2010 FIFA World Cup, Paul the Octopus correctly predicted the outcome of 8 out of 8 matches (all the games played by Germany (went to semis), + the finals). What are the chances of this happening by random chance?

$$
\Pr(\text{success}) = 0.5 \equiv p
$$

$$
\Pr(\text{2 successes}) = p^2 = 0.25
$$

$$
\Pr(\text{8 successes}) = p^8 \approx 0.0039
$$

Paul seems to be a genius. The issue: you too could be in the animal fortune-telling business if you had one animal. In fact, if you had a pet shop, you could turn all your animals into fortune tellers. Let's see how the probability of having one genius pet changes as you increase the number of pets. Let $p_G = 0.0039$ be the probability of a pet being a genius. Then, with $n$ pets, the probability of no genius pet is $(1-p_G)^n$. So the probability of at least one genius pet is:

$$
P_G = 1 - (1 - p_G)^n
$$

```{r}
#| label: prob-genius

p_genius <- 0.5^8
n_pets <- 1:1000
pr_1_genius <- 1 - (1 - p_genius)^n_pets

tibble(
  n = n_pets,
  p = pr_1_genius
) |>
  ggplot(aes(x = n, y = p)) +
  geom_line() +
  geom_hline(yintercept = .5, color = "red") +
  scale_x_continuous(breaks = seq(0, 1000, 200)) +
  labs(
    x = "Number of pets",
    y = "Pr(at least 1 genius)"
  )
```

## How to cheat at regression

That logic extends to p-values. Let's say you are 95% certain that your regression coefficient is zero. That means that if you were to repeat the experiment many times with similar data, 

- 95% of the time you would get a p-value greater than 0.05 $\Rightarrow$ not significant 
- 5% of the time you would get a p-value less than 0.05 $\Rightarrow$ significant. 

So, **the more tests you run, the more likely it is that you'll get a significant result by random chance.**

This means that

- If you run a regression with many variables, you are more likely to find a significant coefficient by random chance.
- If you run many regressions, you are more likely to find a significant coefficient by random chance.

### How to do it in practice? Data mining

1. Pick a dataset with many variables, and don't have a clear hypothesis in mind. 
2. Run many regressions with different combinations of variables until you find a significant coefficient. 
3. Build a nice story about significant coefficient you found, and pretend this was your idea all along.

You can also do this with experiments. Typically, running many treatments is expensive. Collecting data on many outcomes, however, is cheap. So you can run one experiment, and then test many outcomes until you find a significant result.

## How to avoid it?

1. **Statistical adjustments.** There are ways to adjust p-values for multiple testing. The most common one is the Bonferroni correction, which divides the significance level by the number of tests. For example, if you run 10 tests, you would use a significance level of 0.005 instead of 0.05. This makes it harder to find significant results by random chance.
2. **Pre-registration.** This is the practice of specifying your hypotheses and analysis plan before collecting data. You deposit your plan in an archive (e.g., https://osf.io). This way, you can't change your hypotheses after seeing the data. This is becoming very common in the social sciences.

## Another thing that feels like cheating, but may not be

We have seen that regression is bad at predicting non-linear relationships. So, if you have a non-linear relationship, you can try to transform your variables to make the relationship linear. For example, if you have data that is highly skewed (like income), it is often useful to run the following regression:

$$
\log(\text{income}) = \beta_0 + \beta_1 \text{education} + \epsilon
$$

Is this cheating? Individually, no. It is a common practice to transform variables to make them more suitable for regression. However, if you try many different transformations until you find one that gives you a significant result, then you may be back to the problem of multiple testing. 

```{r}
#| label: log-transformation

# we generate some data where the true relationship is linear
x <- rnorm(1000)
y <- 3 * x + 5 + rnorm(1000)
# but the data that we get is exponentiated (like income)
ey <- exp(y)

df <- tibble(
  x = x,
  y = y,
  ey = ey
)

# when we look at the data we observe, the relationship is weak
# because we have a lot of outliers on the right
ggplot(df, aes(x = x, y = ey)) +
  geom_point(alpha = .2) +
  geom_smooth(method = "lm")

# when we use the log, we restore the true relationship
ggplot(df, aes(x = x, y = log(ey))) +
  geom_point(alpha = .2) +
  geom_smooth(method = "lm")
```


## Asides

1. Missing data
    - We'll cover this next Monday
    - Standard solution: drop observations with missing data (R does it automatically)
    - Problem 1: sample size drops $\Rightarrow$ less likely to get significant relationships
    - Problem 2: sample selection. 
    - How to assess? Check whether coefficients on other variables change drastically (i.e., much larger/smaller, or change sign). 
        - No: no selection
        - Yes: selection
2. Standardized variables
    - Definition: consider variable $x$ with mean $\mu$ and standard deviation $\sigma$. The standardized variable $x_{\text{std}} = \frac{x - \mu}{\sigma}$
    - Why does it matter? 
        1. Standardization helps remove differences in the data. Example: you want to compare students across schools. School A grades severely: there's an average of 5/10, and a SD of 2. School B is more lenient: the average is 8/10 and the SD is 1. Thus, a student that got 9/10 is somehow "better" in school A than in school B. This will be reflected in the standardized scores (9/10 = 2 in school A, 1 in school B).
        2. When data is normally distributed, one third of the data is between -1 SD and the mean, and 1 third is between the mean and +1 SD. So, increasing a variable by 1 SD makes you move from the median to the upper third. Increasing a variable by 2 SD makes you move from the bottom third to the upper third (see plot below). 
        3. Thus, we can use standardized scores to compare across quantities that would otherwise be incomparable. Say you're studying the determinants of educational performance. You find that income and study time both increase educational performance. Specifically, you find that increasing income by $100 increases test scores by .2, and increasing study time by one hour increases test scores by .1. Which of these variables has the largest effect? It is unclear how to compare one hour to $100. Considering standardized variables solves the issue, as you can speak of increasing income and study time by 1 SD. 

```{r}
#| label: normal distribution

v <- rnorm(1e4, mean = 0, sd = 1)
ggplot(tibble(data = v), aes(data)) +
  geom_density() +
  geom_vline(xintercept = c(-1, 1), color = "red") +
  geom_vline(xintercept = 0) +
  scale_x_continuous(breaks = seq(-4, 4, 1)) +
  labs(
    title = "A normal distribution",
    caption = "2/3 of the data is between the 2 red lines"
  )
```