---
title: "Lecture 18-19: Webscrapping"
---

```{r}
#| label: setup

library(tidyverse)
library(httr2) # make http requests and handle responses
library(jsonlite) # convert JSON to R objects
library(rvest) # HTML parsing: extract elements from a page and convert to R objects
```

Web scraping gives you access to data at a scale that would be impossible to collect by hand. A few years ago, I scraped the results of every municipal election in Morocco — roughly 235,000 records spread across 1,300 towns and their sub-districts — in a matter of days. Doing the same work manually with a team of research assistants would have taken months and introduced far more error. But the project also taught me what happens when you scrape carelessly: the server eventually detected the volume of requests, geo-blocked access from outside Morocco, and added anti-scraping protections. The data is still there; we just can't get it the same way anymore. The technical skills below are straightforward; the judgment about how to use them responsibly matters just as much.

## How does the web work?

The internet has become one of the richest data sources available to social scientists. News archives, government documents, film databases, social media posts, song catalogues — nearly anything published online is, in principle, accessible to quantitative analysis. **Web scraping** is the practice of extracting that data programmatically: writing code to send requests to web servers and parsing the responses into structured data we can analyze in R.

To scrape effectively, you need to understand the basic mechanics of web communication. Every interaction with the web is a request-response exchange between a **client** (your browser, or your R script) and a **server**.

### Requests

A request has four components:

- **URL** — The address of the resource. It identifies the server (`https://itunes.apple.com`) and the specific resource on that server (`/search`). Optional *query parameters*, appended after a `?`, let you filter or configure the request — for example, `?term=beyonce&media=music&limit=50`.
- **Method** — The operation you want to perform. The two you will encounter:
  - `GET`: retrieve data from the server. This is what your browser does every time you visit a page, and what we will use almost exclusively when scraping.
  - `POST`: send data *to* the server (e.g., submitting a login form or uploading a file). Some APIs require POST for write operations, but most read-only APIs use GET.
  TODO: alude to other methods (PUT, DELETE, etc.) used in REST APIs, but we won't cover them in this course
- **Headers** — Metadata attached to the request, invisible to ordinary users. Common headers include `User-Agent` (identifying which software is making the request) and `Cookie` (session credentials for authenticated sites).
- **Body** — A payload sent alongside the request. Mostly used with POST to transmit form data or structured data. GET requests typically have no body.

### Responses

The server's reply has three parts:

- **Status code** — A three-digit number summarizing what happened.
  - `200 OK`: success.
  TODO: add details
  - `4xx` codes mean *you* made an error: `400` (malformed request), `401` (authentication required), `403` (forbidden), `404` (not found).
  - `5xx` codes mean the *server* has a problem: `500` (internal server error), `503` (service unavailable).
- **Headers** — Metadata about the response: content type (`text/html`, `application/json`), caching instructions, rate-limit counters.
- **Body** — The actual content — an HTML page, a structured data object, an image, a CSV file — whatever was requested.

## Webscraping with APIs

Many websites expose a structured data interface called an **Application Programming Interface (API)**. Rather than returning a full HTML page designed for browsers, an API returns data in a machine-readable format, almost always **JavaScript Object Notation (JSON)** — a plain-text structure built from nested key-value pairs.

```json
{
  "resultCount": 2,
  "results": [
    {
      "artistName": "Beyoncé",
      "trackName": "Crazy in Love",
      "releaseDate": "2003-06-24T07:00:00Z",
      "trackTimeMillis": 235373,
      "primaryGenreName": "R&B/Soul"
    },
    {
      "artistName": "Beyoncé",
      "trackName": "Halo",
      "releaseDate": "2008-11-18T08:00:00Z",
      "trackTimeMillis": 261493,
      "primaryGenreName": "R&B/Soul"
    }
  ]
}
```

When an API is available, prefer it over scraping raw HTML. APIs are more stable (they do not change when the site redesigns its visual layout), more efficient (you get only the data you need), and more legitimate (you are using the intended data access method).

### Example: the iTunes Search API

Apple provides a free, no-credentials-required search API for its music and media catalogue. Let's use it to ask: **have popular songs gotten shorter over time?**

We'll search for a handful of well-known artists with long careers and compare average track duration across the years.

#### Step 1: Send a request

```{r}
#| label: itunes-request

resp <- request("https://itunes.apple.com/search") |>
  req_url_query(
    term = "beyonce",
    media = "music",
    entity = "song",
    limit = 200
  ) |>
  req_perform()

resp
```

`request()` builds a request object. `req_url_query()` attaches query parameters as key-value pairs — no manual URL construction needed. `req_perform()` sends the request and returns the response.

How do we know which endpoints and query parameters to use? The API documentation is the best source. For iTunes, see [the official docs](https://performance-partners.apple.com/search-api). The `entity = "song"` parameter tells the API to return individual tracks; without it you would also get albums, music videos, and other media types mixed into the results.

#### Step 2: Inspect the response

```{r}
#| label: itunes-status

resp |> resp_status()
```

A status of 200 means success. Before you do anything else with the response, consider saving the raw body to disk:

```r
write_lines(resp |> resp_body_string(), "data/raw_response.json")
```

Requests are the expensive part — they count against rate limits, cost money on pay-per-call APIs, and risk triggering a ban if you repeat them. Parsing is cheap and you can redo it as many times as you like. Storing the raw response locally means a bug in your parsing code does not force you to re-hit the server.

Let's look at the response body:

```{r}
#| label: itunes-inspect

# not run because the body is long and messy, but you can run this to see the raw JSON text
# resp |> resp_body_string()
```

The body is a named list with two elements: `resultCount` (how many records were returned) and `results` (a list where each element is one track). Each track is itself a list of fields — we saw the field names in the JSON example above.

#### Step 3: Parse into a tibble

```{r}
#| label: itunes-parse

df <- fromJSON(resp |> resp_body_string())$results |>
  as_tibble()
df
```


#### Step 4: Clean and analyze

```{r}
#| label: itunes-clean

df |>
  mutate(
    year = as.integer(substr(releaseDate, 1, 4)),
    duration = trackTimeMillis / 1000 / 60
  ) |>
  select(year, duration) |>
  group_by(year) |>
  summarize(duration = mean(duration)) |>
  ggplot(aes(year, duration)) +
  geom_line() +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Average track duration over time",
    x = "Year",
    y = "Duration (minutes)"
  )

```

#### Step 5: Scale up to multiple artists

One artist is suggestive; let's see whether the pattern holds more broadly by repeating the search for several artists and combining the results.

```{r}
#| label: itunes-multi

artists <- c(
  "beyonce",
  "taylor swift",
  "drake",
  "adele",
  "ed sheeran",
  "miles davis",
  "john coltrane",
  "bob dylan",
  "madonna",
  "prince"
)

```

```{r}
#| label: itunes-multi-plot

# let's reuse the code from before, but wrap it in a function that takes an artist name as input
# and returns a cleaned tibble of track durations by year

itunes_req <- function(artist) {
  Sys.sleep(0.5) # be polite and avoid rate limits
  request("https://itunes.apple.com/search") |>
    req_url_query(
      term = artist,
      media = "music",
      entity = "song",
      limit = 200
    ) |>
    req_perform() |>
    resp_body_string() |>
    fromJSON() |>
    pluck("results") |>
    as_tibble() |>
    mutate(
      artist = artist,
      year = as.integer(substr(releaseDate, 1, 4)),
      duration = trackTimeMillis / 1000 / 60
    ) |>
    select(artist, year, duration)
}

# CAREFUL: this will send 10 requests to the iTunes API, which may trigger rate limits if you run it repeatedly
# make sure you only run this once, and consider saving the result to disk for future use

pl <- map_dfr(artists, itunes_req)

pl |>
  # data cleaning: filter out years before the artist's debut
  filter(!(artist == "adele" & year < 1995)) |>
  filter(!(artist == "madonna" & year < 1975)) |>
  filter(!(artist == "prince" & year < 1950)) |>
  filter(!(artist == "prince" & year > 2016)) |>
  filter(!(artist == "miles davis" & year > 1991)) |>
  group_by(year, artist) |>
  summarize(duration = mean(duration)) |>
  ggplot(aes(year, duration)) +
  geom_line() +
  geom_point() +
  facet_wrap(~artist, ncol = 3, scale = "free") +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Average track duration over time",
    x = "Year",
    y = "Duration (minutes)"
  )

```

## Webscraping with HTML

Not every website offers an API. Often the data we want lives directly in a webpage — encoded in the **HyperText Markup Language (HTML)** that browsers parse and render visually. We can still extract it programmatically by navigating the HTML structure.

HTML is a tree of nested **elements**. Each element has a tag name, optional attributes, and content. A simple table looks like this:

```html
<table>
  <tr>
    <th>Title</th>     <th>Year</th>  <th>Gross</th>
  </tr>
  <tr>
    <td>Avengers: Endgame</td>  <td>2019</td>  <td>$2,799,439,100</td>
  </tr>
</table>
```

To extract specific elements from a page, we use **CSS selectors** — patterns that match elements by their tag, class, or identifier:

| Selector | Matches |
|---|---|
| `table` | Any `<table>` element |
| `.wikitable` | Any element with `class="wikitable"` |
| `#summary` | The element with `id="summary"` |
| `table.wikitable` | A `<table>` element with class `wikitable` |

The `rvest` package provides a clean interface for fetching pages and navigating their structure.

### Finding the right selector

Before writing any scraping code you need to know which selector to use. Browser developer tools are the fastest way to find out.

**Opening DevTools**: right-click any element on the page and choose *Inspect*, or press Shift-Ctrl-I (Windows/Linux) or Cmd-Option-I (Mac). This opens a panel showing the live HTML of the page.

**Inspector tab**: hover over elements in the HTML tree and the corresponding element highlights on the page (and vice versa). This lets you confirm you are targeting the right node. Once you have found it, right-click the element in the Inspector → *Copy* → *Copy selector* to get a CSS selector automatically. Chrome's generated selectors are often overly specific (long chains of tags and pseudo-class indices); simplify them — `table.wikitable` is more robust than `body > div#content > table:nth-child(3)`.

**Network tab**: every HTTP request the page makes is listed here, including background requests for data. If the content you want does not appear in the raw HTML (a common sign that JavaScript is rendering it), check the Network tab — you may find that the page is actually calling an internal API, in which case you can call that API directly instead of scraping HTML.

### Example: Wikipedia's highest-grossing films

Wikipedia's list of highest-grossing films is published as a `wikitable` — a consistently-structured HTML table that parses cleanly. Let's scrape it and ask: **has the box-office ceiling risen over time?**

#### Step 1: Fetch and parse the page

```{r}
#| label: wiki-fetch

page <- read_html(
  "https://en.wikipedia.org/wiki/List_of_highest-grossing_films"
)
page
```

`read_html()` fetches the page over HTTP and parses the HTML into a tree we can navigate. The output tells us how many nodes the document contains — we don't read those directly; we use selectors instead.

#### Step 2: Find the table

```{r}
#| label: wiki-table

films_raw <- page |>
  html_element("table.wikitable") |>
  html_table()

films_raw
```

`html_element()` finds the first element matching the selector. `html_table()` converts a `<table>` node directly into a tibble, handling the row-and-column structure for us.

#### Step 3: Clean the data

```{r}
#| label: wiki-clean

films <- films_raw |>
  select(title = `Title`, year = `Year`, gross = `Worldwide gross`) |>
  mutate(
    gross = parse_number(gross)
  )
```



#### Step 4: Analyze

```{r}
#| label: wiki-plot

ggplot(films, aes(x = year, y = gross)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(
    labels = scales::dollar_format(scale = 1e-9, suffix = "bn")
  ) +
  labs(
    title = "Box-office ceiling over time",
    x = "Year",
    y = "Worldwide gross (USD)"
  )
```

This comparison is not inflation-adjusted, so it is biased toward more recent films — a dollar in 1994 is worth about twice a dollar today. Correcting for inflation would require merging in a historical price index; the scraping is the easy part.

## Webscraping with a headless browser

The two approaches above cover most scraping scenarios, but both break down on one type of site: pages where the content is **rendered by JavaScript** in the browser rather than included in the initial HTML. Modern single-page applications, live dashboards, and some search results fall into this category. If you fetch a page and find that the table visible in your browser is absent from the raw HTML, JavaScript rendering is the likely cause.

The solution is a **headless browser** — a full web browser that runs without a visible window, executes JavaScript, and lets you extract the fully-rendered result. The main R option is `chromote`. We will not cover it in this course, but the [`chromote` documentation](https://rstudio.github.io/chromote/) is a good starting point.

## Good practices for webscraping

### Don't overload the server

Scraping in a loop — requesting hundreds of pages in rapid succession — can look like a denial-of-service attack to the receiving server. Most servers enforce rate limits and will block your IP address if you exceed them. Even if they do not, sending a flood of requests is impolite and can degrade the experience for other users.

**Add delays between requests.** The simplest approach is `Sys.sleep()` between loop iterations, as we used in the iTunes example. For more precise control, `httr2` provides `req_throttle()`, which enforces a maximum request rate across all requests to a given host:

```{r}
#| eval: false

request("https://example.com/api") |>
  req_throttle(rate = 10 / 60) # at most 10 requests per minute
```

### Identify yourself

By default, `httr2` sends a generic user agent string. Many server administrators appreciate knowing who is making requests — and some services block unidentified or suspicious agents. Set a descriptive `User-Agent` header:

```{r}
#| eval: false

request("https://example.com") |>
  req_headers(`User-Agent` = "ResearchProject/1.0 (contact@university.edu)")
```

### Write robust code

Networks are unreliable. A scraper that crashes on the first failed request is useless for bulk data collection. Two tools help:

**`req_retry()`** automatically retries a request after transient failures — status 429 (too many requests), 503 (service unavailable), or network errors — with optional exponential backoff:

```{r}
#| eval: false

request("https://example.com") |>
  req_retry(max_tries = 3, backoff = function(i) {
    2^i
  })
```

The `backoff` function controls the wait between attempts: with `2^i`, the first retry waits 2 seconds, the second 4 seconds, the third 8 seconds. This gives the server time to recover rather than hammering it in quick succession.

Note that not all failures are worth retrying. Status 404 (not found) is a permanent failure — the resource does not exist and retrying will not help. Status 429 and 503 are temporary — the server is busy or rate-limiting you, and a short wait often resolves them. `req_retry()` handles this distinction automatically by default.

**`purrr::possibly()`** wraps any function so that instead of throwing an error on failure, it returns a default value (typically `NULL`). This lets a loop continue even when individual requests fail. The philosophy: some errors you *want* to crash on — a bug in your own code, for instance — but others you want to absorb and move on from, like a single bad URL in a list of 500. `possibly()` is the tool for the second category.

```{r}
#| eval: false

addition <- function(a, b) {
  a + b
}

safe_addition <- possibly(addition, otherwise = NULL)

addition(1, 2)
addition(1, "a")
safe_addition(1, 2)
safe_addition(1, "a")

safe_scrape <- possibly(
  function(url) {
    read_html(url) |> html_element("table") |> html_table()
  },
  otherwise = NULL
)

results <- map(urls, safe_scrape) |> compact() # compact() drops the NULLs
```

`possibly()` is a convenience wrapper around R's base `tryCatch()`, which gives you finer-grained control — you can handle errors, warnings, and messages separately, or inspect the error message to decide what to do. For most scraping purposes `possibly()` is enough, but `tryCatch()` is useful when you want to log the specific error or take different actions on different failure types:

```{r}
#| eval: false

safe_scrape_verbose <- function(url) {
  tryCatch(
    {
      read_html(url) |> html_element("table") |> html_table()
    },
    error = function(e) {
      message("Failed on ", url, ": ", conditionMessage(e))
      NULL
    }
  )
}
```

### Your scraper should be able to be interrupted and restarted

A well-written scraper can be interrupted and restarted without repeating work that already succeeded. This matters because scraping jobs are often long-running — a single run may take hours — and failures (network timeouts, rate-limit bans, power outages) are common. The solution is three habits used together: **save results as you go**, **log success and failure**, and **skip items you already have**.

Think of it as a two-column spreadsheet with a URL column and a status column (initially empty). Before each request you check the status: if it is already "ok", skip it. After each request you write the result to disk and update the status to "ok" or "failed" — immediately, not at the end of the loop. If the script crashes halfway through, every completed URL is already recorded and the next run will pick up exactly where the last one left off.

```{r}
#| eval: false

urls <- c(
  "https://example.com/data1",
  "https://example.com/data2",
  "https://example.com/data3",
  "https://example.com/data4",
  "https://example.com/data5"
)

dir.create("data/raw", showWarnings = FALSE)
log_file <- "data/scrape_log.csv"

# Initialize log if it doesn't exist yet
if (!file.exists(log_file)) {
  tibble(url = urls, status = as.characer(NA)) |> write_csv(log_file)
}

log <- read_csv(log_file, show_col_types = FALSE)

for (url in urls) {
  # Skip if already succeeded
  if (log$status[log$url == url] == "ok") {
    next
  }

  # use a safe scraping function to make sure that
  # 1. you retry a few times upon failure
  # 2. potential errors don't crash the entire loop
  # -> instead, you just move on to the next URL
  result <- safe_scrape(url) # returns NULL on failure

  if (is.null(result)) {
    log$status[log$url == url] <- "failed"
  } else {
    out_file <- paste0("data/raw/", URLencode(url, repeated = TRUE), ".csv")
    write_csv(result, out_file)
    log$status[log$url == url] <- "ok"
  }

  write_csv(log, log_file) # persist log after every iteration
  Sys.sleep(0.5)
}

# Load and combine all saved results
results <- list.files("data/raw", pattern = "\\.csv$", full.names = TRUE) |>
  map_dfr(read_csv, show_col_types = FALSE)
```

The important point is that the log and the data are written to disk after every single iteration, not at the end of the loop. If the script crashes on iteration 3, iterations 1–2 are already saved and logged; restarting will skip them and resume from iteration 3.

### Cache your results locally

Once a scraper has run successfully and the data is on disk, protect it from accidental re-runs. The simplest guard is a file existence check at the top of the data-collection block:

```{r}
#| eval: false

results_file <- "data/results.csv"

if (!file.exists(results_file)) {
  # run the scraper and save results
  results <- map(urls, safe_scrape) |> compact() |> bind_rows()
  write_csv(results, results_file)
}

results <- read_csv(results_file, show_col_types = FALSE)
```

This way, re-running or re-rendering your document reads from the saved file instead of re-issuing requests. If you genuinely need fresh data, delete the file and re-run.
