---
title: "Lecture 17: Missing data and measurement error"
---

- Today, we will discuss problems with data quality: measurement error and missing data.
- Two big categories: 
  - Quality problems are random (e.g. random measurement error, missing completely at random)
  - Quality problems are systematic (e.g. systematic measurement error, missing not at random)

Which one is worse? 

Systematic problems are worse, but random problems are still a problem.

## Measurement error

###

- Statistics is based on the idea that there are errors in our data, and helps us to understand how to deal with them.
- It's this idea that we observe $\tilde{y} = y + \epsilon$, where $\epsilon$ is the measurement/sampling error.  
  - $y$ is the true temperature of the water, $\tilde{y}$ is the observed temperature measured by a real (imperfect) thermometer. $\epsilon$ is the *measurement* error in our measurement.
  - $y$ is average height of students in a class, $\tilde{y}$ is the observed height of one student, and $\epsilon$ is the *sampling* error.

###
```{r}
#| label: measurement-error

## a plot with vertical bar at 100C water temperature + normal PDF around it
```

###

- taking the average of many measurements $\tilde{y}_1, \tilde{y}_2, \ldots, \tilde{y}_n$ cancels out errors cancel
  - We can recover the true value $y$
  - and derive confidence intervals etc.
- This only works because the errors are random.

### Problems with random measurement error

Experiment: 

1. at what temperature does water boil? 
  - I want to measure $y$ (boiling point of water)
  - $y = \tilde{y} + \epsilon$
2. how does that vary with salinity? 
  - I want to measure $y$ (boiling point of water) as a function of $x$ (salinity)
  - $y = \alpha + \beta x + \epsilon$

  ###

- Answering question 1: my measure will be noisy, but more measurements/better measurement will help (so, small problem).
- Answering question 2: 
  - measurement error in $y$: my measure of the relationship between boiling point and salinity ($\beta$) will be noisy, but more measurements/better measurement will help (so, small problem).
  - measurement error in $x$: my measure of the relationship between boiling point and salinity ($\beta$) will be biased towards zero (attenuation bias; only better measurement helps (so, big problem)).

### Systematic measurement error

- Random measurement error: some thermometers are off by some temperature (above, below): $y = \tilde{y} + \epsilon$
- Systematic measurement error: all thermometers from one supplier are off by 1 degree above: $y = \tilde{y}$ + 1 + \epsilon$
- Why is this a problem? 
  - We will overestimate the boiling point of water (even if those thermometers are randomly distributed across experimenters).
  - There is no way to fix this problem with more measurements (we will just get more measurements of the wrong value).

### In the social sciences: random or systematic measurement error?

- On a scale of 1 to 10, how happy are you? 
  - Random measurement error: some people are in a bad mood, some people are in a good mood, and that affects their answer.
- Have you ever paid a bribe?
  - Systematic measurement error: people are likely to underreport this, so we will underestimate the true value.

### Solutions to systematic measurement error

1. Use better measurement tools (e.g. better survey questions, better thermometers).
2. Think of the direction of the bias and interpret results accordingly. Example: 
  - Question 1: do you support Donald Trump? (pre 2024) 
  - Question 2: trust in government?
  - Finding: government trust among Trump supporters = 0.3, government trust among non-Trump supporters = 0.5.
  - Interpretation: Trump supporters are less trusting of government than non-Trump supporters, but the true difference is likely larger than 0.2 (because non-Trump supporters contain liars, who, if they behave like honest Trump supporters, have lower government Trump).

## Missing data

### 

- We can think of missing data as a measurement error where the error is so large that we can't even observe the value (i.e. it's missing).
- The same issues carry over
  - If data are missing at random, the sample is smaller, which increases the noise (though here, we don't have attenuation bias).
  - If data are missing systematically, the sample is biased, which can lead to incorrect conclusions.

### Solutions

1. Throw away the data (i.e., only analyze observations where all variables are present).
2. Treat missing data as a separate category (e.g. create a dummy variable for missingness).
  - Good test for whether data are missing at random or not: if the missingness dummy variable is correlated with the outcome variable, then data are likely missing systematically.
2. Impute the missing data 
  - Simple: assign the mean (does not help when data systematically missing systematically)
  - Complicated: using a regression model to predict the missing values based on other variables.
  - Imputation can help to reduce noise
    - Example: an observation has 1/10 variables missing. It's a shame to throw it away. Use a regression model to predict the missing variable based on the other 9 variables. This means we can use this observation in our analysis, benefiting from the 9 observed variables. 
    - That's provided our imputation procedure works well (and that's a big if, there's often no way to know).

## In practice

###

- We often have a mix of random and systematic measurement error, and a mix of missing data that is random and systematic.
- And we're not 100% sure which of our control variables play a causal role and which do not
- So we usually show results
  - with and without controls
  - keeping sample size constant (i.e. only using observations with no missing data)

### A bad example

```{r}
use GSS data, show a regression of income on education, then income on education + race + age
```

### Good example 1

```{r}
use GSS data, show a regression of income on education, then income on education + race + age, using the same sample (i.e. only observations with no missing data)
```

### Good example 2

```{r}
use GSS data, show a regression of income on education (full sample), then income on education (small sample), then income on education + race + age, using the same sample (i.e. only observations with no missing data)
```

