---
title: "Lecture 16: Data visualization"
---

```{r}
#| label: setup
#| echo: false

suppressPackageStartupMessages(library(tidyverse))
library(gapminder)
```

## Principles of good data visualization

Good data visualization is not just about aesthetics. It is about **communicating information as clearly and honestly as possible**. The most influential framework for thinking about this comes from Edward Tufte, a statistician and designer who laid out a set of principles in his 1983 book *The Visual Display of Quantitative Information*.

The five core principles are:

1. **Data-ink ratio.** Every drop of ink should earn its place by encoding data. Remove everything that does not carry information: background colors, redundant gridlines, unnecessary borders, decorative elements.

2. **Graphical integrity.** The visual representation must not distort the data. The most common violation is truncating the y-axis to make a small difference look large.

3. **Data density.** Show as much data as possible in a compact space. Aggregating to a single mean hides variation; showing the raw distribution is almost always better.

4. **Small multiples.** When a relationship varies across groups, repeat the same chart structure for each group side by side. This makes comparisons natural and avoids overloaded single plots.

5. **Direct labeling.** Label data directly on the chart rather than forcing the reader to look up a legend. Every time a reader's eyes travel away from the data, information is lost.

These principles are not rules to follow mechanically. They are heuristics that help you ask: *is every element of this plot earning its place?*

```{mermaid}
graph TD
  A[Data-ink ratio] --> D[Data density]
  B[Graphical integrity] --> D
  D --> E[Small multiples]
  D --> F[Direct labeling]
  style A fill:#4a90d9,color:#fff
  style B fill:#4a90d9,color:#fff
  style D fill:#27ae60,color:#fff
  style E fill:#f39c12,color:#fff
  style F fill:#f39c12,color:#fff
```

---

## Example 1: Data-ink ratio

We use the `mpg` dataset (built into ggplot2), which contains fuel economy data for 234 cars across 7 vehicle classes.

**A first attempt.** Here is a bar chart of mean highway fuel efficiency by vehicle class.

```{r}
#| label: mpg-bad

mpg |>
  group_by(class) |>
  summarize(mean_hwy = mean(hwy)) |>
  ggplot(aes(x = class, y = mean_hwy, fill = class)) +
  geom_col() +
  labs(
    title = "Highway fuel efficiency by vehicle class",
    x = "Vehicle class",
    y = "Mean highway mpg",
    fill = "Class"
  )
```

**What is wrong with this plot?**

**A better version.**

## Example 2: Graphical integrity

We use the resume experiment dataset from Bertrand and Mullainathan (2004). Researchers sent fictitious resumes with stereotypically Black or White names to job postings and recorded whether the applicant received a callback.

We start from a regression model. Regressing callback on race gives us predicted callback rates for each group, along with standard errors we can use to construct confidence intervals.

```{r}
#| label: resume-setup
#| echo: false

df_resume <- read_csv("./data/lecture-13-resume.csv", show_col_types = FALSE)
```

```{r}
#| label: resume-model

# Regress callback on race
mod <- lm(call ~ race, data = df_resume)

# Predict callback rates for each group at 90% and 95% confidence levels
preds <- tibble(race = c("black", "white"))
ci <- predict(mod, preds, interval = "confidence", level = 0.95)

# Combine into one tibble
preds <- bind_cols(preds, as_tibble(ci))
```

**A misleading plot.**

```{r}
#| label: resume-bad

ggplot(preds, aes(x = race, y = fit, fill = race, ymin = lwr, ymax = upr)) +
  geom_col() +
  geom_errorbar() +
  labs(
    title = "Racial discrimination in hiring: a massive gap",
    x = "Applicant race",
    y = "Callback rate",
    fill = "Race"
  )
```

**What is wrong with this plot?**

**A better version.**

---

## Example 3: Small multiples and data density

We use the `gapminder` dataset, which tracks life expectancy, GDP per capita, and population for 142 countries across 5 continents from 1952 to 2007.

**The spaghetti plot.** Here is life expectancy over time for all countries at once.

```{r}
#| label: gapminder-bad

ggplot(gapminder, aes(x = year, y = lifeExp, color = country)) +
  geom_line() +
  theme(legend.position = "right") +
  labs(
    title = "Life expectancy over time",
    x = "Year",
    y = "Life expectancy (years)"
  )
```

**What is wrong with this plot?**

**A better version.** 

---

## Example 4: Direct labeling

We use the `txhousing` dataset (built into ggplot2), which tracks monthly median house sale prices across Texas cities. We compare 6 major cities.

```{r}
#| label: txhousing-setup
#| echo: false

cities <- c(
  "Houston",
  "Dallas",
  "Austin",
  "San Antonio",
  "Fort Worth",
  "El Paso"
)

tx_annual <- txhousing |>
  filter(city %in% cities) |>
  group_by(city, year) |>
  summarize(median_price = median(median, na.rm = TRUE), .groups = "drop")
```

**A plot with a legend.**

```{r}
#| label: tx-bad

ggplot(tx_annual, aes(x = year, y = median_price, color = city)) +
  geom_line() +
  labs(
    title = "Median house prices in Texas cities",
    x = "Year",
    y = "Median sale price ($)",
    color = "City"
  )
```

**What is wrong with this plot?**

- To identify a line, the reader must: (1) look at the line, (2) note its color, (3) scan the legend, (4) return to the data. This happens every time.
- The default ggplot2 color palette is not optimized for perceptual distinctiveness.
- The legend occupies space outside the plot, reducing the data area.

**Direct labeling: place the city name at the end of each line.**

```{r}
#| label: tx-good

# Keep only the last observation per city to position the labels
tx_labels <- tx_annual |>
  group_by(city) |>
  slice_max(year, n = 1)

ggplot(tx_annual, aes(x = year, y = median_price, color = city)) +
  geom_line(linewidth = 0.8) +
  # place label just to the right of the terminal point
  geom_text(
    data = tx_labels,
    aes(label = city),
    hjust = 0, # anchor text to the left (so it extends rightward)
    nudge_x = 0.3,
    size = 3
  ) +
  # extend the x-axis to make room for labels on the right
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.2))) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Median house prices in Texas cities",
    x = "Year",
    y = "Median sale price ($)"
  )
```

The reader can now follow any line directly to its label without leaving the data area. Key changes: `geom_text()` with `hjust = 0` places city names just to the right of each line's endpoint; `scale_color_brewer(palette = "Dark2")` provides a more perceptually distinct palette; `scale_x_continuous(expand = ...)` creates space for the labels on the right.
